---
title: "Assignment 1"
author: "Cory J. Cascalheira"
date: "1/26/2021"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assignment Instructions

In a Word document, respond to the following and upload via Canvas:

1. Download the rollercoasters.sav file and open it in SPSS (or import it into R)
2. Write up a response to the following question, using an ANOVA: Do roller coasters that have different excitement ratings differ in their max speed?
    * Make sure to describe center, spread, and shape of the distribution of the variables (Analyze, Descriptives, Explore may be helpful here)
    * Are assumptions of ANOVA met? (What are the assumptions? You can look in your book for this, or also use Laerd Statistics (Links to an external site.))
    * Include all relevant output from the ANOVA in your response to the question

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
